Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-23 src/1.Introduction.tex
Further further further (this one was my fault)

Parallel and distributed computing moved into focus at universities and teaching institutions. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations. At Sandia, we develop a performance portability library called Kokkos that is ideal for educators looking to teach parallel computing.

In teaching a class on parallel programming, educators must balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do many programming models cleanly express all parallel programming constructs. Many vendor models compound this, requiring students have access to specific hardware, and limiting the concepts of parallelism to those supported by that architecture. This makes the choice of programming model important and difficult.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine model specification. We would like to compare this to the support of Von Neumann architectures in programming languages today. We define an abstract machine model later in this work. The lack of such parallel programming models motivated us to develop Kokkos.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world's biggest supercomputers. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes functions representing parallel patterns and types for data management and maps them to the underlying abstract execution model. This execution model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}. Its key properties are

\begin{itemize}
\item Parallel programming model for C++
\item Targets an abstract machine model
\item Implements parallel patterns and tasking paradigms
\item Offers abstract data types for performance portability
\item Pure C++ implementation through template metaprogramming
\item Aligns with C++ standardization efforts for parallel programming
\item Vendor- and architecture independent and open-source
\item Includes exercises and teaching material
\end{itemize}

@@

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators. We show an abstraction with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

>>>>

Parallel and distributed computing is a focus at universities and teaching institutions. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula. Secondly, parallel computing is an attractive field for department research and industry collaborations. At Sandia, we develop a performance portability library called Kokkos that is ideal for educators looking to teach parallel computing.

In teaching a class on parallel programming, educators must balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do many programming models cleanly express all parallel programming constructs. Many vendor models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware. This makes the choice of programming model important and difficult.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a programming exercise for students becomes capable of running efficiently on parallel computer architectures that conform to a common parallel execution model. This compares to the support of Von Neumann architectures in programming languages today. The lack of such parallel programming models motivated us to develop Kokkos.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world's biggest supercomputers. It consists of a programming model specification and a library. The specification defines an execution and memory model and an application programming interface (library API). The runtime library exposes functions representing parallel patterns and types for data management and maps them to the underlying abstract execution model. The Kokkos library maps those patterns to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}. Its key properties are

\begin{itemize}
\item Parallel programming model for C++
\item Targets an abstract execution model
\item Implements parallel patterns and tasking paradigms
\item Offers abstract data types for performance portability
\item Pure C++ implementation through template metaprogramming
\item Aligns with C++ standardization efforts for parallel programming
\item Vendor- and architecture independent and open-source
\item Includes exercises and teaching material
\end{itemize}

@@

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators. We show an abstraction with enough semantic information to produce performance-portable, parallel code on current architectures. We highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
